An Empirical Analysis of the Role of Energy in Economic Growth

### 2013-04-30 EconGrowth-R-Functions.Rnw
More editing and fixing the graphs. Still have a problem with stars appearing in the gamma grid graph key.
Changed the size of symbols in the gamma grid search 9-panel lattice graph.
Now using country names (instead of abbreviations) in partial residual plots.

# Statistical significance levels. We'll work with 95% CIs
ciLevel <- 0.95
ciHalfLevel <- ciLevel + (1.0-ciLevel)/2.0
# List of countries
countryAbbrevs <- c(US="US", UK="UK", JP="JP", CN="CN", ZA="ZA", SA="SA", IR="IR", TZ="TZ", ZM="ZM")
countryAbbrevsAlph <- c(CN="CN", IR="IR", JP="JP", SA="SA", TZ="TZ", UK="UK", US="US", ZA="ZA", ZM="ZM")
countryAbbrevsU <- c(US="US", UK="UK", JP="JP") #Only these countries have useful work data
countryAbbrevsAlphU <- c(JP="JP", UK="UK", US="US")
countryNamesAlph <- c(CN="China", IR="Iran", JP="Japan", SA="Saudi Arabia", TZ="Tanzania", UK="United Kingdom", US="USA", ZA="South Africa", ZM="Zambia") #In alphabetical order.
yLimitsForGDPGraphs <- list(c(1,10), c(1,4), c(1,4), c(1,4), c(1,4), c(1,4), c(1,4), c(1,4), c(1,4)) # Alph order
energyTypes <- c(Q="Q", X="X", U="U") # List of energy types
factors <- c(K="K", L="L", Q="Q", X="X", U="U") # List of factors of production
########### Several global parameters for graphs. Set here and use below to ensure consistent appearance of graphs.
# Set the order for presenting countries in 3x3 lattice graphs. Default is alphabetical. 
# "1" means first alphabetically.
countryOrderForGraphs <- c(7,6,3,1,8,5,4,2,9) # Sets the order as US, UK, JP, CN, ZA, TZ, SA, IR, ZM.
countryOrderForGraphsU <- c(3, 2, 1) # Sets the order as US, UK, JP when we're looking only at energyType=U.
timeTics <- c(1980, 1990, 2000, 2010)
yTicsForIndexedGraphs <- c(1,2,3,4,5,6,7,8,9,10) #y tic mark locations.

maxWidth <- 6.5 #Inches
# Full page lattice plot sizes
ninePanelLayoutSpec <- c(3,3) #indicates a 3x3 arrangement of panels.
threePanelLayoutSpec <- c(3,1) #indicates a 1 row x 3 column arangement of panels
onePanelLayoutSpec <- c(1,1) #indicates a 1x1 arrangement of panels.
ninePanelGraphWidth <- maxWidth     
ninePanelGraphHeight <- 7.5  #Inches
presentationGraph1RowHeight <- 2.8 #Inches
presentationGraph2ColWidth <- 2 #Inches
presentationGraph1ColWidth <- 2*presentationGraph2ColWidth
keyTextSize <- 0.85 #85% of normal size
keyColumns <- 1 #Want only 1 column in the key for lattice graphs
defaultKeyXLoc <- 0.01 #x position of the key. This default is good for 9-panel graphs.
defaultKeyYLoc <- 0.95 #y position of the key. This default is good for 9-panel graphs.
# Parameter graphs are tricky to ensure same size. I've found that with height set to 5 in for the 
# cd graphs and 2.93 in for the sf graphs, I get a nearly-exact match with graph size. From these settings I 
# can set up 2 equations:
#    5 in = 2*hgraph + haxis
#    2.93 in = hgraph + haxis
# to estimate that h_axis = 0.86 in and hgraph = 2.05 in.
h_param_graph <- 2.07 # Inches for a single panel
h_param_axis <- 0.86 # Inches (also includes whatever margin is applied)
sfParameterGraphHeight <- h_param_graph + h_param_axis
cdParameterGraphHeight <- 2*h_param_graph + h_param_axis
cesParameterGraphHeight <- 3*h_param_graph + h_param_axis
linexParameterGraphHeight <- h_param_graph + h_param_axis
sfParameterGraphWidth <- maxWidth
cdParameterGraphWidth <- maxWidth
cesParameterGraphWidth <- maxWidth
linexParameterGraphWidth <- maxWidth
# Other graph parameters that apply to all graphs
scaleTextSize <- 0.8  #Multiple of normal size
scaleTickSize <- -0.5 #50% of normal size and pointing INWARD!
@

@@

  symbols <- c(NA, NA, NA, 1)

@@

@@

@@

                     pch = c(1), col.symbol = c("black"), #Controls symbol parameters

@@

#################
# Creates a plot with y residuals vs. energy (e). The y residuals are from the specified modelType without energy.
# modelType: one of "CD" or "CES"
# energyType: one of "Q", "X", or "U" to serve as the ordinate of the graph
# countryAbbrev: set to NA (the default) if you want a lattice graph with all countries shown. Set to one of
#                US, UK, JP, CN, ZA, SA, IR, TZ, ZM for an individual graph.
## 
createPartialResidualPlot <- function(modelType, energyType, countryAbbrev=NA, textScaling=1.0){
  xLabelString <- paste("$", tolower(energyType), "$", sep="")
  yLabelString <- paste("$y$ residuals (", modelType, " without energy)", sep="")
  if (! is.na(countryAbbrev)){
    # An individual graph for a single country is desired.
    data <- calcResiduals(modelType=modelType, countryAbbrev=countryAbbrev, energyType=energyType)
    plot <- xyplot(resid ~ iEToFit, data=data,
                   type=c("p"),
                   col="black",
                   scales=list(cex=scaleTextSize*textScaling, #controls text size on scales.
                               tck=scaleTickSize), #controls tick mark length. < 0 for inside the graph.
                   xlab=list(label=xLabelString, cex=textScaling),
                   ylab=list(label=yLabelString, cex=textScaling)
                   )
  } else {
    # Lattice graph will all countries is desired.
    data <- do.call("rbind", lapply(countryAbbrevsAlph, calcResiduals, modelType=modelType, energyType=energyType))
    # If useful work is desired, need to trim the data set.
    if (energyType == "U"){
      # Use data from only those countries that have U defined.
      data <- subset(data, !is.na(iEToFit))
print(data)
      layout <- threePanelLayoutSpec
      factorLevels <- countryAbbrevsAlphU # Want only those countries for which U is known
print(factorLevels)
      indexCond <- list(countryOrderForGraphsU) # Orders as US, UK, JP
    } else {
      # Use data from all countries, because all countries have Q and X data.
      factorLevels <- countryNamesAlph # We want all countries shown
      layout <- ninePanelLayoutSpec
      indexCond <- list(countryOrderForGraphs) # Orders as US, UK, JP, CN, ZA, SA, IR, TZ, ZM
    }
    plot <- xyplot(resid ~ iEToFit | Country, data=data,
                   type = c("p"),
                   index.cond = indexCond, #orders the panels.
                   layout = layout,
                   strip = strip.custom(factor.levels=factorLevels, 
                                        bg="transparent", # Sets background transparent to match the graph itself.
                                        par.strip.text=list(cex=textScaling) # Scales text in the strip.
                                        ),
                   as.table = TRUE, #indexing of panels starts in upper left and goes across rows.
                   col = "black",
                   scales = list(cex=scaleTextSize*textScaling, #controls text size on scales.
                                 tck=scaleTickSize, #controls tick mark length. < 0 for inside the graph.
                                 alternating=FALSE), # prevents left/right, top/bot alternating
                   xlab = list(label=xLabelString, cex=textScaling),
                   ylab = list(label=yLabelString, cex=textScaling)
                   )
  }
  return(plot)
}

>>>>

# Statistical significance levels. We'll work with 95% CIs
ciLevel <- 0.95
ciHalfLevel <- ciLevel + (1.0-ciLevel)/2.0
# List of countries
countryAbbrevs <- c(US="US", UK="UK", JP="JP", CN="CN", ZA="ZA", SA="SA", IR="IR", TZ="TZ", ZM="ZM")
countryAbbrevsAlph <- c(CN="CN", IR="IR", JP="JP", SA="SA", TZ="TZ", UK="UK", US="US", ZA="ZA", ZM="ZM")
countryAbbrevsU <- c(US="US", UK="UK", JP="JP") #Only these countries have useful work data
countryAbbrevsAlphU <- c(JP="JP", UK="UK", US="US")
countryNamesAlph <- c(CN="China", IR="Iran", JP="Japan", SA="Saudi Arabia", TZ="Tanzania", UK="United Kingdom", US="USA", ZA="South Africa", ZM="Zambia") #In alphabetical order.
countryNamesAlphU <- c(JP="Japan", UK="United Kingdom", US="USA") #In alphabetical order.
yLimitsForGDPGraphs <- list(c(1,10), c(1,4), c(1,4), c(1,4), c(1,4), c(1,4), c(1,4), c(1,4), c(1,4)) # Alph order
energyTypes <- c(Q="Q", X="X", U="U") # List of energy types
factors <- c(K="K", L="L", Q="Q", X="X", U="U") # List of factors of production
########### Several global parameters for graphs. Set here and use below to ensure consistent appearance of graphs.
# Set the order for presenting countries in 3x3 lattice graphs. Default is alphabetical. 
# "1" means first alphabetically.
countryOrderForGraphs <- c(7,6,3,1,8,5,4,2,9) # Sets the order as US, UK, JP, CN, ZA, TZ, SA, IR, ZM.
countryOrderForGraphsU <- c(3, 2, 1) # Sets the order as US, UK, JP when we're looking only at energyType=U.
timeTics <- c(1980, 1990, 2000, 2010)
yTicsForIndexedGraphs <- c(1,2,3,4,5,6,7,8,9,10) #y tic mark locations.

maxWidth <- 6.5 #Inches
# Full page lattice plot sizes
ninePanelLayoutSpec <- c(3,3) #indicates a 3x3 arrangement of panels.
threePanelLayoutSpec <- c(3,1) #indicates a 1 row x 3 column arangement of panels
onePanelLayoutSpec <- c(1,1) #indicates a 1x1 arrangement of panels.
threePanelGraphWidth <- maxWidth
threePanelGraphHeight <- 3.17 #Inches
ninePanelGraphWidth <- maxWidth     
ninePanelGraphHeight <- 7.5  #Inches
presentationGraph1RowHeight <- 2.8 #Inches
presentationGraph2ColWidth <- 2 #Inches
presentationGraph1ColWidth <- 2*presentationGraph2ColWidth
keyTextSize <- 0.85 #85% of normal size
keyColumns <- 1 #Want only 1 column in the key for lattice graphs
defaultKeyXLoc <- 0.01 #x position of the key. This default is good for 9-panel graphs.
defaultKeyYLoc <- 0.95 #y position of the key. This default is good for 9-panel graphs.
# Parameter graphs are tricky to ensure same size. I've found that with height set to 5 in for the 
# cd graphs and 2.93 in for the sf graphs, I get a nearly-exact match with graph size. From these settings I 
# can set up 2 equations:
#    5 in = 2*hgraph + haxis
#    2.93 in = hgraph + haxis
# to estimate that h_axis = 0.86 in and hgraph = 2.05 in.
h_param_graph <- 2.07 # Inches for a single panel
h_param_axis <- 0.86 # Inches (also includes whatever margin is applied)
sfParameterGraphHeight <- h_param_graph + h_param_axis
cdParameterGraphHeight <- 2*h_param_graph + h_param_axis
cesParameterGraphHeight <- 3*h_param_graph + h_param_axis
linexParameterGraphHeight <- h_param_graph + h_param_axis
sfParameterGraphWidth <- maxWidth
cdParameterGraphWidth <- maxWidth
cesParameterGraphWidth <- maxWidth
linexParameterGraphWidth <- maxWidth
# Other graph parameters that apply to all graphs
scaleTextSize <- 0.8  #Multiple of normal size
scaleTickSize <- -0.5 #50% of normal size and pointing INWARD!
@

@@

@@

  symbolSize <- 0.5
  symbolType <- 0 #for an open square
  symbols <- c(NA, NA, NA, symbolType)

@@

                              lines=list(lty=lineTypes, lwd=lineWidths, pch=symbols, cex=symbolSize), #controls lines
                              col=colors,

@@

                     pch = c(symbolType), col.symbol = c("black"), cex = symbolSize, #Controls symbol parameters

@@

#################
# Creates a plot with y residuals vs. energy (e). The y residuals are from the specified modelType without energy.
# modelType: one of "CD" or "CES"
# energyType: one of "Q", "X", or "U" to serve as the ordinate of the graph
# countryAbbrev: set to NA (the default) if you want a lattice graph with all countries shown. Set to one of
#                US, UK, JP, CN, ZA, SA, IR, TZ, ZM for an individual graph.
## 
createPartialResidualPlot <- function(modelType, energyType, countryAbbrev=NA, textScaling=1.0){
  xLabelString <- paste("$", tolower(energyType), "$", sep="")
  yLabelString <- paste("$y$ residuals (", modelType, " without energy)", sep="")
  if (! is.na(countryAbbrev)){
    # An individual graph for a single country is desired.
    data <- calcResiduals(modelType=modelType, countryAbbrev=countryAbbrev, energyType=energyType)
    plot <- xyplot(resid ~ iEToFit, data=data,
                   type=c("p"),
                   col="black",
                   scales=list(cex=scaleTextSize*textScaling, #controls text size on scales.
                               tck=scaleTickSize), #controls tick mark length. < 0 for inside the graph.
                   xlab=list(label=xLabelString, cex=textScaling),
                   ylab=list(label=yLabelString, cex=textScaling)
                   )
  } else {
    # Lattice graph will all countries is desired.
    data <- do.call("rbind", lapply(countryAbbrevsAlph, calcResiduals, modelType=modelType, energyType=energyType))
    # If useful work is desired, need to trim the data set.
    if (energyType == "U"){
      # Use data from only those countries that have U defined.
      data <- subset(data, !is.na(iEToFit))
      layout <- threePanelLayoutSpec
      factorLevels <- countryNamesAlphU # Want only those countries for which U is known
      indexCond <- list(countryOrderForGraphsU) # Orders as US, UK, JP
    } else {
      # Use data from all countries, because all countries have Q and X data.
      factorLevels <- countryNamesAlph # We want all countries shown
      layout <- ninePanelLayoutSpec
      indexCond <- list(countryOrderForGraphs) # Orders as US, UK, JP, CN, ZA, SA, IR, TZ, ZM
    }
    plot <- xyplot(resid ~ iEToFit | Country, data=data,
                   type = c("p"),
                   index.cond = indexCond, #orders the panels.
                   layout = layout,
                   strip = strip.custom(factor.levels=factorLevels, 
                                        bg="transparent", # Sets background transparent to match the graph itself.
                                        par.strip.text=list(cex=textScaling) # Scales text in the strip.
                                        ),
                   as.table = TRUE, #indexing of panels starts in upper left and goes across rows.
                   col = "black",
                   scales = list(cex=scaleTextSize*textScaling, #controls text size on scales.
                                 tck=scaleTickSize, #controls tick mark length. < 0 for inside the graph.
                                 alternating=FALSE), # prevents left/right, top/bot alternating
                   xlab = list(label=xLabelString, cex=textScaling),
                   ylab = list(label=yLabelString, cex=textScaling)
                   )
  }
  return(plot)
}

### 2013-04-30 EconGrowth.Rnw
Now using country names (instead of abbreviations) in partial residual plots.

\item{Heun: Adjust variable threePanelGraphHeight in the Functions file to produce U partial residual plots that are exactly the same size as the graphs in the 9-panel plots.}

@@

@@

<<CDeGridLattice, fig.width=ninePanelGraphWidth, fig.height=ninePanelGraphHeight, fig.cap="Grid search over $\\gamma$ for Cobb-Douglas with energy model.", eval=TRUE>>=
cdeGridLattice(energyType="Q")
@

@@

<<CESPartialResidualsX, fig.width=ninePanelGraphWidth, fig.height=ninePanelGraphHeight, fig.align="center", fig.cap="CES partial residuals plot for exergy ($x$).">>=
createPartialResidualPlot(modelType="CES", energyType="X", countryAbbrev=NA)
@

<<CESPartialResidualsU, fig.width=threePanelGraphWidth, fig.height=threePanelGraphHeight, fig.align="center", fig.cap="CES partial residuals plot for useful work ($u$).">>=
createPartialResidualPlot(modelType="CES", energyType="U", countryAbbrev=NA)
@

>>>>

\item{Heun: Ensure consistent x and y axis limits in partial residual plots}

@@

\item{\st{Heun: Use full country names instead of 2-letter abbreviations in partial residual plots}}
\item{\st{Heun: Adjust variable threePanelGraphHeight in the Functions file to produce U partial residual plots that are exactly the same size as the graphs in the 9-panel plots.}}

@@

<<CDeGridLattice, fig.width=ninePanelGraphWidth, fig.height=ninePanelGraphHeight, fig.cap="Grid search over $\\gamma$ for Cobb-Douglas with energy model.">>=
cdeGridLattice(energyType="Q")
@

@@

<<CESPartialResidualsX, fig.width=ninePanelGraphWidth, fig.height=ninePanelGraphHeight, fig.align="center", fig.cap="CES partial residuals plot for exergy ($x$).", eval=TRUE>>=
createPartialResidualPlot(modelType="CES", energyType="X", countryAbbrev=NA)
@

<<CESPartialResidualsU, fig.width=threePanelGraphWidth, fig.height=threePanelGraphHeight, fig.align="center", fig.cap="CES partial residuals plot for useful work ($u$).", eval=TRUE>>=
createPartialResidualPlot(modelType="CES", energyType="U", countryAbbrev=NA)
@
